Gaussian Process regression for astronomical time-series

### 2022-09-14 src/tex/ms.tex
[showyourwork] overleaf sync

The most widely-used algorithms for scalable exact GPR in time domain astrophysics are based---often not explicitly---on stochastic differential equations \citep[SDEs;][]{Sarkka:2019}, and typically require that the data be ``sortable''.
As an early example in the astrophysics literature, \citet{pr95} presented an algorithm for GPR with linear scaling with the size of the dataset, when using an exponential kernel with unevenly sampled time series data.
This method was later generalized to a wider range of covariance functions \citep{ambikasaran2015generalized, celerite, 2020AJ....160..240G}, and the so-called \project{celerite}\footnote{\url{https://github.com/dfm/celerite}} implementation has been widely-used for time domain astronomy \dan{Add some citations to show the range?} \suz{yes. I take it you're reluctant to cite the S+LEAF papers?}.
A qualitatively different, and independently developed approach to implementing these models uses a state-space representation for the underlying SDE to enable scalable inference \citep{2014ApJ...788...33K,  2021RNAAS...5..107J, 2022arXiv220709327M} \dan{Maybe add some citations of the use of these methods}.

@@

One area of active development concerns the problem of non-Gaussianity. While the assumption of Gaussianity implicit to GPR is often good enough in practice, it is almost never strictly correct, and this can become problematic in some circumstances, such as when extreme precision is required on the model hyper-parameters, or when the data contain (a relatively small proportion of) significant outliers. In particular, Student's-$t$ processes could be an interesting solution for the former case, and Dirichlet processes for the latter.

, which also share the analytic marginalisation properties of GPs (although the equations are not as simple). These could prove interesting in some applications, though it is unclear whether they would deliver better performance than a GP on data with outliers remains to be demonstrated.

\suz{FUTURE DEVELOPMENTS -- talk about continued use in areas where it's already pretty widespread like RVs, but also new time-domain surveys like Rubin-LSST, SKA, transient astronomy. Talk about the development of more physically motivated GP models e.g. in asteroseismology and in making inferences about stellar and planetary surfaces (c.f. starry-process) -- any ideas for where else this might be possible. Talk about other classes of stochastic processes maybe for cases when non-Gaussianity is important (but I really don't know much about this!)
Talk about new kinds of GP models bridging the gap between "traditional models and AI, like Gaussian Process networks [ref recent RV paper]}

>>>>

The most widely-used algorithms for scalable exact GPR in time domain astrophysics are based---often not explicitly---on stochastic differential equations \citep[SDEs;][]{Sarkka:2019}, and typically require that the data be ``sortable''.
As an early example in the astrophysics literature, \citet{pr95} presented an algorithm for GPR with linear scaling with the size of the dataset, when using an exponential kernel with unevenly sampled time series data.
This method was later generalized to a wider range of covariance functions \citep{ambikasaran2015generalized, celerite, 2020AJ....160..240G, 2020A&A...638A..95D, 2022A&A...659A.182D}, and the \project{celerite}\footnote{\url{https://github.com/dfm/celerite}} and \project{S+LEAF}\footnote{\url{https://obswww.unige.ch/~delisle/spleaf/doc}} implementations have been widely-used for time domain astronomy. \dan{Add some citations to show the range.}
% \suz{yes. I take it you're reluctant to cite the S+LEAF papers?}.
A qualitatively different, and independently developed approach to implementing these models uses a state-space representation for the underlying SDE to enable scalable inference \citep{2014ApJ...788...33K,  2021RNAAS...5..107J, 2022arXiv220709327M} \dan{Maybe add some citations of the use of these methods}.

@@

One area of active development concerns the problem of non-Gaussianity. While the assumption of Gaussianity implicit to GPR is often good enough in practice, it is almost never strictly correct, and this can become problematic when extreme precision is required on the model hyper-parameters, or when the data contain (a relatively small proportion of) significant outliers. Stochastic process models based on distributions with higher kurtosis than the Gaussian, for example Student's-$t$ processes \citep[see e.g.][]{Tracey_2018} are an interesting alternative, though they have not been extensively used in astronomy to our knowledge.

Another area where GPR could become more widely used in future is for intelligent observation planning: if a set of variable objects is being monitored as part of a survey, using a GP to model existing observations would enable the observer to make predictions (including uncertainties) for future observations, and therefore to prioritise among the target list in real time according to some metric of choice. \suz{Keep or remove? I'm not sure we have anything truly meaningful to say here.}

\suz{It would be nice to include a few words about new kinds of GP models bridging the gap between "traditional models" and AI, like Gaussian Process networks [ref recent RV paper] - but I am not sure I know enough about them.}
